---
title: "Cyclistic Data Analysis"
author: "usafhulk"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Package to handle this large csv file

```{r}
#package to handle this large csv file
options(repos = c(CRAN = "[https://cran.rstudio.com](https://cran.rstudio.com)")) # Explicitly set mirror here
#package to handle this large csv file
install.packages("data.table")
library(data.table)
```

# Reading in all the csv files for the previous 12 months of data

```{r}
Divvy_24_Feb <- fread("202402-divvy-tripdata.csv")
Divvy_24_Mar <- fread("202403-divvy-tripdata.csv")
Divvy_24_Apr <- fread("202404-divvy-tripdata.csv")
Divvy_24_May <- fread("202405-divvy-tripdata.csv")
Divvy_24_Jun <- fread("202406-divvy-tripdata.csv")
Divvy_24_Jul <- fread("202407-divvy-tripdata.csv")
Divvy_24_Aug <- fread("202408-divvy-tripdata.csv")
Divvy_24_Sep <- fread("202409-divvy-tripdata.csv")
Divvy_24_Oct <- fread("202410-divvy-tripdata.csv")
Divvy_24_Nov <- fread("202411-divvy-tripdata.csv")
Divvy_24_Dec <- fread("202412-divvy-tripdata.csv")
Divvy_25_Jan <- fread("202501-divvy-tripdata.csv")
Divvy_25_Feb <- fread("202502-divvy-tripdata.csv")
```

# Efficient loop that verifies the colnames of each data.table without typing out each individual name.

```{r}
months <- c("Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
years <- c(rep(24, 11), 25, 25) # 24 for Feb-Dec, 25 for Jan-Feb

for (i in 1:length(months)) {
  df_name <- paste0("Divvy_", years[i], "_", months[i])
  if (exists(df_name)) { # Check if the data frame exists
    print(colnames(get(df_name)))
  } else {
    print(paste("Data frame", df_name, "does not exist."))
  }
}
```

# Combining all the dataframes into one dataframe

```{r}
Divvy_24_25 <- rbind(Divvy_24_Feb, Divvy_24_Mar, Divvy_24_Apr, Divvy_24_May, Divvy_24_Jun, Divvy_24_Jul, Divvy_24_Aug, Divvy_24_Sep, Divvy_24_Oct, Divvy_24_Nov, Divvy_24_Dec, Divvy_25_Jan, Divvy_25_Feb)
```

# Inspecting the data

```{r}
head(Divvy_24_25)
str(Divvy_24_25)
summary(Divvy_24_25)
missing_coords <- which(is.na(Divvy_24_25$end_lat))
summary(missing_coords)
```

# Create a smaller subset (e.g., 10,000 rows)

```{r}
subset_size <- 10000
subset_dt <- Divvy_24_25[1:subset_size]
```

# This data.table is large, so i am using a subset join method for this process. I am looking to fix the NA values in the end_lat and end_lng columns. To do this, I am searching for the first non-NA value in the end_lat and end_lng columns for each end_station_name. This allows me to find the correct end_lat and end_lng values for the missing rows and complete the data.table. Running this as a loop would take far too much memory and computation time, this is why i went with the join method. 
```{r}
# Measure start time
start_time <- Sys.time()

# Create valid coordinates data.table (using the full data.table)
valid_coords <- Divvy_24_25[!is.na(end_lat), .(end_lat = end_lat[1], end_lng = end_lng[1]), by = end_station_name]

# Update the main data.table using a join (using the full data.table)
Divvy_24_25[is.na(end_lat), `:=`(end_lat = valid_coords[.SD, end_lat, on = "end_station_name"],
                                 end_lng = valid_coords[.SD, end_lng, on = "end_station_name"])]

# Measure end time
end_time <- Sys.time()

# Calculate elapsed time
elapsed_time <- end_time - start_time

# Print elapsed time
print(paste("Elapsed time:", elapsed_time))
```

# Verify the changes

```{r}
summary(Divvy_24_25)
missing_coords <- which(is.na(Divvy_24_25$end_lat))
summary(missing_coords)
```

# Check for missing values

```{r}
missing_values <- colSums(is.na(Divvy_24_25))
print(missing_values)
```

# Check for duplicated rows

```{r}
duplicated_rows <- sum(duplicated(Divvy_24_25))
print(duplicated_rows)
```

# Write to a new CSV file using fwrite to efficiently handle large data in the data.table

```{r}
#fwrite(Divvy_24_25, "Divvy_24_25.csv")
summary(Divvy_24_25)
tail(Divvy_24_25, 20)
```

# Calculate trip duration in minutes

```{r}
Divvy_24_25[, trip_duration := as.numeric(difftime(ended_at, started_at, units = "mins"))]
```

# Visualize the distribution of trip duration for members and casual riders

```{r}
library(ggplot2)

# Visualize the usage patterns over time
ggplot(Divvy_24_25, aes(x = started_at, fill = member_casual)) +
  geom_histogram(binwidth = 3600, position = "dodge") +
  labs(title = "Usage Patterns Over Time", x = "Start Time", y = "Count") +
  theme_minimal()
```

# Calculate the number of trips per user type

```{r}
trip_counts <- Divvy_24_25[, .N, by = .(member_casual)]
print(trip_counts)
```

# Calculate average trip duration per user type

```{r}
avg_trip_duration <- Divvy_24_25[, .(avg_duration = mean(trip_duration, na.rm = TRUE)), by = .(member_casual)]
print(avg_trip_duration)
```

# Identify peak usage times

```{r}
Divvy_24_25[, hour_of_day := hour(started_at)]
peak_usage <- Divvy_24_25[, .N, by = .(member_casual, hour_of_day)]
print(peak_usage)
```

# Visualize trip counts by user type

```{r}
ggplot(trip_counts, aes(x = member_casual, y = N, fill = member_casual)) +
  geom_bar(stat = "identity") +
  labs(title = "Trip Counts by User Type", x = "User Type", y = "Count") +
  theme_minimal()
```

# Visualize average trip duration by user type

```{r}
ggplot(avg_trip_duration, aes(x = member_casual, y = avg_duration, fill = member_casual)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Trip Duration by User Type", x = "User Type", y = "Average Trip Duration (minutes)") +
  theme_minimal()
```

# Visualize peak usage times

```{r}
ggplot(peak_usage, aes(x = hour_of_day, y = N, color = member_casual, group = member_casual)) +
  geom_line() +
  labs(title = "Peak Usage Times by User Type", x = "Hour of Day", y = "Count") +
  theme_minimal()
```

# Perform a t-test to compare trip durations between members and casual riders

```{r}
t_test_result <- t.test(trip_duration ~ member_casual, data = Divvy_24_25)
print(t_test_result)
```

# Function to calculate distance using the Haversine formula

```{r}
haversine <- function(lat1, lon1, lat2, lon2) {
  R <- 3959 # Earth radius in miles
  delta_lat <- (lat2 - lat1) * pi / 180
  delta_lon <- (lon2 - lon1) * pi / 180
  a <- sin(delta_lat / 2) * sin(delta_lat / 2) +
    cos(lat1 * pi / 180) * cos(lat2 * pi / 180) *
    sin(delta_lon / 2) * sin(delta_lon / 2)
  c <- 2 * atan2(sqrt(a), sqrt(1 - a))
  R * c
}

# Calculate distance for each trip
Divvy_24_25[, distance := haversine(start_lat, start_lng, end_lat, end_lng)]
```

# Aggregate the data by binning trip duration and distance

```{r}
library(dplyr)
binned_data <- Divvy_24_25 %>%
  mutate(
    trip_duration_bin = cut(trip_duration, breaks = seq(0, max(trip_duration), by = 5)),
    distance_bin = cut(distance, breaks = seq(0, max(distance), by = 0.5))
  ) %>%
  group_by(member_casual, trip_duration_bin, distance_bin) %>%
  summarise(
    avg_trip_duration = mean(trip_duration, na.rm = TRUE),
    avg_distance = mean(distance, na.rm = TRUE),
    count = n()
  ) %>%
  ungroup()
```

# Calculate the 10th and 90th percentiles for trip duration and distance

```{r}
trip_duration_cutoffs <- quantile(Divvy_24_25$trip_duration, probs = c(0.1, 0.9), na.rm = TRUE)
distance_cutoffs <- quantile(Divvy_24_25$distance, probs = c(0.1, 0.9), na.rm = TRUE)
```

# Filter the data to remove the top and bottom 10%

```{r}
filtered_data <- Divvy_24_25 %>%
  filter(trip_duration >= trip_duration_cutoffs[1], trip_duration <= trip_duration_cutoffs[2],
         distance >= distance_cutoffs[1], distance <= distance_cutoffs[2])
```

# Sample 10% of the filtered data

```{r}
set.seed(123) # for reproducibility
sampled_data <- filtered_data %>% sample_frac(0.1)
```

# Create a box plot to visualize the distribution of trip durations by user type

```{r}
ggplot(sampled_data, aes(x = member_casual, y = trip_duration, fill = member_casual)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Trip Durations by User Type",
    x = "User Type",
    y = "Trip Duration (minutes)",
    fill = "User Type"
  ) +
  scale_fill_manual(values = c("member" = "blue", "casual" = "red")) +
  theme_minimal()
```

# Create a histogram to show the distribution of trip durations by user type

```{r}
ggplot(sampled_data, aes(x = trip_duration, fill = member_casual)) +
  geom_histogram(binwidth = 5, position = "dodge") +
  labs(
    title = "Distribution of Trip Durations by User Type",
    x = "Trip Duration (minutes)",
    y = "Count",
    fill = "User Type"
  ) +
  scale_fill_manual(values = c("member" = "blue", "casual" = "red")) +
  theme_minimal()
```

# Create a density plot to show the distribution of trip durations by user type

```{r}
ggplot(sampled_data, aes(x = trip_duration, fill = member_casual)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density of Trip Durations by User Type",
    x = "Trip Duration (minutes)",
    y = "Density",
    fill = "User Type"
  ) +
  scale_fill_manual(values = c("member" = "blue", "casual" = "red")) +
  theme_minimal()
```

# Analyze start station usage

```{r}
start_station_usage <- Divvy_24_25[, .N, by = .(member_casual, start_station_name)]

# Find the top 10 start stations for each user type
top_start_stations <- start_station_usage[ , head(.SD, 10), by = member_casual][order(-N), ]

print("Top 10 Start Stations by User Type:")
print(top_start_stations)
```

# Visualize top start stations

```{r}
ggplot(top_start_stations, aes(x = start_station_name, y = N, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 10 Start Stations by User Type",
       x = "Start Station",
       y = "Number of Trips") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate labels
```

# Analyze end station usage (similar to start station analysis)

```{r}
end_station_usage <- Divvy_24_25[, .N, by = .(member_casual, end_station_name)]

# Find the top 10 end stations for each user type
top_end_stations <- end_station_usage[ , head(.SD, 10), by = member_casual][order(-N), ]

print("Top 10 End Stations by User Type:")
print(top_end_stations)
```

# Visualize top end stations

```{r}
ggplot(top_end_stations, aes(x = end_station_name, y = N, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 10 End Stations by User Type",
       x = "End Station",
       y = "Number of Trips") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
library(lubridate) # Make sure lubridate is loaded

# Add a column to indicate the day of the week
Divvy_24_25[, day_of_week := wday(started_at, abbr = TRUE)] # Use abbr instead of label

# Analyze trip counts by day of week and user type
usage_by_day <- Divvy_24_25[, .N, by = .(member_casual, day_of_week)]

print("Usage by Day of Week:")
print(usage_by_day)
```

# Visualize usage by day of week

```{r}
ggplot(usage_by_day, aes(x = day_of_week, y = N, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Usage by Day of Week",
       x = "Day of Week",
       y = "Number of Trips") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
library(lubridate) 

# Add a column to indicate the day of the week
Divvy_24_25[, day_of_week := wday(started_at, abbr = TRUE)] # Use abbr instead of label

# Analyze trip counts by day of week and user type
usage_by_day <- Divvy_24_25[, .N, by = .(member_casual, day_of_week)]

print("Usage by Day of Week:")
print(usage_by_day)
```

# Visualize usage by day of week 

```{r}
ggplot(usage_by_day, aes(x = day_of_week, y = N, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Usage by Day of Week",
       x = "Day of Week",
       y = "Number of Trips") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Add a column to indicate the month

```{r}
Divvy_24_25[, month := month(started_at, abbr = TRUE)] # Use abbr instead of label
```

# Analyze trip counts by month and user type

```{r}
usage_by_month <- Divvy_24_25[, .N, by = .(member_casual, month)]

print("Usage by Month:")
print(usage_by_month)
```

# Visualize usage by month 

```{r}
ggplot(usage_by_month, aes(x = month, y = N, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Usage by Month",
       x = "Month",
       y = "Number of Trips") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```